---
title: "Breeze_HW_5"
author: "Kelley Breeze"
date: "2022-09-20"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Tasks

### 1. Read in the horseshoe crab data set and modify the variables as listed below. Print the updated data object out.

Reading in the horseshoe crab data set and saving as an R object called `crabs`
```{r}
crabs<-read_table('crabs.txt')
```


Modify the `color`, `spine condition` and `y` variables as listed below. Remove column of `NA` values.
1.  `color` is a factor where:  
+   2=`light`  
+   3=`medium`  
+   4=`dark`  
+   5=`darker`  
2.  `spine condition` is a factor where:  
+   1=`both good`  
+   2=`one worn or broken`  
+   3=`both worn or broken`  
3.  `y` is a factor where:  
+   0=`No Satellite`  
+   1=`At least 1 Sattelite`  

```{r}
crabs<- crabs%>%
  mutate(
    color= factor(color,
                       levels = c(2, 3, 4, 5),
                       labels = c("light",
                                  "medium",
                                  "dark",
                                  "darker")),
    spine = factor(spine,
                   levels = c(1, 2, 3),
                   labels = c("Both Good", "One Worn/Broken", "Both Worn/Broken")),
    y = factor(y,
               levels = c(0, 1),
               labels = c("No Satellite", "At least 1 Satellite")),
) %>%
      select(color:y)
```


### 2. Create a two-way contingency table between the satellite (`y`) and `spine` variables. Write text describing what two of the numbers in the table represent.

The contingency table created here gives us the frequency with which all of the combinations of the `y` variable and `spine` variable occur. The contingency table returns the following information from our data set:  


1.  Number of crabs that have `No` "satellite" and a spine condition of `both good` = 11  
2.  Number of crabs that have `No` "satellite" and `one worn or broken` spine = 8  
3.  Number of crabs that have `No` "satellite" and `both worn or broken` spine = 43  
4.  Number of crabs that have `At` "satellite" and `both good` = 26  
5.  Number of crabs that have `At` "satellite" and `one worn or broken` = 7  
6.  Number of crabs that have `At` "satellite" and `both worn or broken` = 78  

```{r}
satelliteSpine <- table(crabs$y,
                        crabs$spine)
satelliteSpine
```




### 3. Create a three way table between the `color`, `spine`, and satellite (`y`) variables. Output the table. Then, using that table object, print our a two-way table between the spine and satellite for crabs with "darker" color. Write text describing what one of the numbers in the table represents.

Creating a three way table between the `color`, `spine` and `y` variables, saving it as an R object called `colorSpineSatellite` and printing the table.
```{r}
colorSpineSatellite<- table(crabs$color,
                            crabs$spine,
                            crabs$y)
colorSpineSatellite
```

The code below takes the `colorSpineSatillite` object that we just created and prints out a two-way table between the spine and satellite for crabs with "`darker`" `color`. The new contingency table, `spineSatellite`, is a two way contingency table created from the previous three way table `colorSpineSatellite`. In this table we a returning the frequencies of the `spine` and `y` variable combinations for the `darker` level of our `color` variable.  If we look at the `spineSatellite` table we can see that we have 1 crab in the data set that is `darker`, with `No` "satellite" and `both good` for the `spine` condition.
```{r}
spineSatellite<- colorSpineSatellite[4, , ]

spineSatellite
```



### 4. Recreate the side-by-side bar plots below. Write text that comments on any patterns you see in one of the plots.

Code for the first plot, looking at Satellite/No satellite and crab color
```{r}
g1 <- ggplot(data = crabs, aes(x=color))

g1 + geom_bar(aes(fill = y), position = "dodge") + 
  labs(x = "Female Crab Color") + 
  scale_fill_discrete(name=NULL) + 
  coord_flip()  
```

Code for the second plot, looking at Satellite/No satellite and spine condition
```{r}
g2 <- ggplot(data = crabs, aes(x=spine))

g2 + geom_bar(aes(fill = y), position = "dodge") + 
  labs(x = "Female Crab Spine Condition") + 
  scale_fill_discrete(name=NULL) + 
  coord_flip()  
```


Code for the third plot, looking at spine condition and crab color
```{r}
g3 <- ggplot(data = crabs, aes(x=spine))

g3 + geom_bar(aes(fill = color), position = "dodge") + 
  labs(x = "Female Crab Spine Condition") + 
  scale_fill_discrete(name="Female Crab Color") + 
  coord_flip()  
```




### 5. Recreate the set of side-by-side bar plots below. You’ll need to look up how to orient the x-axis lables to 45 degrees. Write text that comments on any patterns you see.

Code for the side-by-side bar plots, looking at crab color and satellite `y` by the spine condition.
```{r}
g5 <- ggplot(data = crabs, aes(x=color))

g5 + geom_bar(aes(fill = y), position = "dodge") + 
  labs(x = "Female Crab Spine Condition") + 
  guides(x=guide_axis(angle=45)) +
  scale_fill_discrete(name=NULL) + 
  facet_grid(cols=vars(spine), labeller = label_both)
```




### 6. Recreate the summary statistics (avg, sd, median, IQR) for the `color` and `y` variables below. Write text that interprets one of the sets of summary stats.

```{r}
crabs%>%
  group_by(color, y)%>%
  summarise(Avg = mean(weight), Sd = sd(weight), Median = median(weight), IQR = IQR(weight))
```

### 7. Recreate the boxplots below showing no satellite and at least 1 sattelite for weight and width. Write text that comments on any patterns you see.

```{r}
g7 <- ggplot(crabs, aes(x = y, y = weight))

weightBox<- g7 + geom_boxplot() + 
  labs(title = "Boxplot for Weight") + 
  guides(fill = legend) + 
  geom_jitter(aes(colour=y))

weightBox
```

```{r}
g7 <- ggplot(crabs, aes(x = y, y = (weight*0.01)))

weightBoxScaled<- g7 + geom_boxplot() + 
  labs(title = "Boxplot for Weight", 
       y = "weight") + 
  geom_jitter(aes(colour=y))

weightBoxScaled
```


### 8. Report the correlation between the weight and width variables. Then recreate the scatterplot below. Write text that comments on any patterns you see.

The correlation between the weight and width variables is 0.899067.
```{r}
Correlation<- cor(select(crabs, weight, width), method = "spearman")

Correlation
```

```{r}
library(ggplot2)
```



Scatterplot showing `weight` and `width` of the crabs, grouped by `color` and `y` variables
```{r}
g8 <- ggplot(crabs, aes(x=weight, y=width))

g8 + geom_point(aes(shape = y, color = color)) +
  geom_smooth(method = lm)
```



